The reproductive and parental care transcriptome of the rock dove

### 2020-09-07 manuscript/elife-v1.tex
now with a table and some results

% !TeX root = RJwrapper.tex
\title{Transcriptional changes in the reproductive axis of parenting pigeons}
\author{by Suzanne H. Austin \footnote{equal contribtion, currently at the
  University of Oregon}, Rayna M Harris \footnote{equal contribtion}, Andrew S. Lang, Victoria S. Farrar, April Booth, Tanner Feustel, Matthew D. MacManes, Rebecca M. Calisi \footnote{corresponding author}}

@@

\abstract{%
In species that provide parental care, successful rearing of offspring
involves a shift from aggressive and sexual behaviors to more caring and
nurturing ones. We hypothesize that these shifts in behavior are related
to changes in gene expression that are controlled by internal mechanisms
rather than by the external cues. We found that gene expression in the
hypothalamus is fairly insensitive to the natural transition between egg
and nestling stages; however, in the pituitary and gonads, gene
expression showed much more plasticity over the course of reproduction.
We show evidence that natural transitions in reproductive and parental
behavior by mediated internal clock controlling gene expression in the
pitary and gonads but that the hypothalamus is very sensitive to
external cues that signal a distruption in the natural parental care
cycle. This research provides a deeper understanding of the interplay of
genes, hormones, and parental care and has important implications for
understanding molecular processes that underlie behavior transitions.
}

@@

Reproduction is a critical period in an organism's life cycle. Parents
must balance their own needs with the energetic costs and
time-constraints required to successfully reproduce and rear their
offspring to independence (cite). A complex series of evolutionarily
conserved physiological processes underlie behaviors typically
associated with reproduction in vertebrates, including courtship,
nesting, and parental care. These processes are primarily mediated by
the reproductive axis, which consists of the hypothalamus in the brain,
the pituitary, and the gonads (i.e., the HPG). Our understanding of the
physiological changes that adults navigate as they transition through
reproductive states are primarily limited to a number of circulating
hormones and a small set of well-known genes (cite). We have yet to come
close to conceptualizing the larger symphony of reproductive events that
occur to yield reproductive and parental care behaviors.

Using the model of a biparental vertebrate species, the rock dove
(Columba livia), our team has previously demonstrated the dynamic nature
of gene activity within the reproductive or HPG axis.
\citet{MacManes2017} reported that HPG gene activity varied by sex, with
females exhibiting higher levels of overall gene expression than males.
\citet{Calisi2018} reported that HPG gene activity varied response to
restraint-stress, and Austin et al.~identify genes that responds to both
corticosterone treatment and restraint stress. While each of these
studies advanced our understanding of sex-differences and
transcriptional plasticity in the reproducitve axis, none of these
studies examined activity of the reproducitve axis of bird that are
actively reproducing or caring for offspring. Studying parental care in
both sexes of mammalian species is challenging because only the females
are able to use lactation to feed their young, and the physiological
mechanisms regulating this process are hard to control for. The rock
dove is an excellent system for studying reproduction and parental care
in both sexes because both males and females feed their young an
energy-rich secretion produced from the epithelial lining of their
crops, known as crop milk.

In this study, we investigate the activity of the HPG axis over the
course of reproduction and parental care by characterizing gene
expression as birds transition from non-breeding to nest-building, as
they lay and incubate their clutches, and finally as they become parents
and raise their chicks. Then, we conducted a series of nest
manipulations to ascertain constraints associated with biological timing
versus input from external stimuli. We conducted a comprehensive series
of analyses examine broad patterns of variation as well as the activity
of specific genes in specific tissues. We have also made our data and
scripts available so that others can reuse the data to ask different
questions or reproduce our results. This research provides a deeper
understanding of the interplay of genes, hormones, and parental care and
has important implications for understanding molecular processes that
underlie behavioral transitions. \citep[reviewed in][]{Lea1989} say
this, but other say that \citep[reviewed in][]{Dulac2014}.

@@

\emph{Data-driven characterization of reproductive and parental stage
groups.}

We first examined broad patterns of variation in gene expression in the
hypothalamus, pituitary, and gonads of male and female rock doves across
a series of timepoints that encapsulate important transitions in
reproduction and and parental care (Fig. 1A). t-Distributed Stochastic
Neighbor Embedding (t-SNE) is a nonlinear, dimension reduction technique
is used to visualize high-dimensional data (van der Maaten and Hinton
2008) and is widely used in in RNA sequencing to identify related and
distinct cell types (refs). Differential expespression profiling is
useful for pair-wise comparisons of differentially expressed genes
(DEGs) with positive log fold change (+LFC) in one group compared to
another. Combined, the provide stronger support for similarities and
differences between cell-types and treatment groups in multi-dimential
datasets. First, the data supports that the hypothalamus, pituitary, and
gonads have distinct gene expression profiles as evidenced by little
overlap in t-SNE space and an average of 13,000 DEGs between each tissue
(Fig. 1B). Second, the data supports that sex-differences are much more
promient in the gonads than in the brain (13,000 versus 2,000 DEGs)
(Fig. 1C). Because of these widespread tissue and sex-differences,
further ananlyses were conducted separately for each tissue and sex
(Fig. 1D, Suppl. Fig 1). In all the tissues, the control samples are
quiet different from all the other timepoints samples, and only a
fraction of the transcriptome is altered during the course of parental
care at the timepoints sampled. The most notable differences male and
female hpothalamic control samples are very different from on the rest
(\textasciitilde{}7000 DEGs, and only the two nestling care stages (n5
and n9)) show differential expression of more than 1,000 genes. In the
pituitary, late inubcation and hatch are quite simlilar to each other
and distinct from the earlier incubation stages stages, and temporal
changes in gene expression appears to be less variable in males than
females. In the female gonads there appear to be more changes in gene
expression early one, between nest building and early incuation, but the
male gonads show very little variation associated with parental stage.

\emph{Hypothesis-driven analysis of candidate parental-care genes.}

\begin{itemize}
\tightlist
\item
  Champagne Curley Genes (Fig 2)
\item
  GO genes (Fig 2 Suppl Fig 1)
\item
  Cancer genes (Fig 2 Suppl Fig 2)
\item
  Shiny genes (Fig 2 Suppl Fig 3)
\end{itemize}

\emph{Software for data exploration and hypothesis testing}

Data accessibility and reproducibility are key challenges for today's
researchers. Using Binder and Shiny, we have provides a means for
scientists, physicians, and the general public to explore the data,
code, and results from this project at
\url{https://raynamharris.shinyapps.io/musicalgenes/}. Like Hipposeq
(Cembrowski et al elife 2016, 2018), this tools allows quick and easy
data exploration of RNAseq data from mutiple cell types that relate to
brain function.

@@

This RNA-sequencing project was designed to characterize the
reproductive axis of bi-parental pigeons across a characteristic
reproductive cycle and through the parental main parental stages of egg
incubation and parental care using data- and hypotheses driven
approaches. We hypothesized that differences in gene expression would be
driven by either internal physiology or external stimuli. In the
pituitary, we found that the most explanatory source of variation in
gene expression was the expression level of prolactin (PRL). Circulating
levels of prolactin in the blood in this species are known to rise
around day nine and peak on or the day before hatch. Our results confirm
this and also identified one hundred genes whose expression is highly
correlated with prolactin. Today, this variation, which changes more
within a parental stage (e.g.~egg incubation or nestling care) suggests
that internal physiology drives changes in gene expression that precede
changes in external stimuli that lead to behavioral changes (e.g.~crop
milk or regurgitation). We predicted that either sex steroid peptide
hormones would play a role in governing the reproductive transitions,
but we did not anticipate discovering a link between PRL expression to
genes that are involved with DNA repair and regulation of the cycle.
Finally, we aimed to make these data and analyses FAIR (findable,
accessible, interoperable, and reproducible), open and reproducible to
encourage others to reuse our data to test new hypotheses or explore the
data with a different perspective or toolset. The scripts that made the
figures and tables are browseable online thanks to RStudio, GitHub and
Binder. The data can be visualized and sonified (turned into sound)
interactively thanks to RStudio, GitHub, and Shiny.

@@

Thanks to Owen for discussion of data sonfification. Thanks to DIB lab
members for software and methods discussion (Titus, Taylor, Luiz,
Amanda, Alicia, Picaso). We thank Pacha for help with database and
Shiny.

We thank Irene, Erica, Beth Krestoff and Tiffany Chen for assistance in
managing the aviary and collecting nest stage data. We thank Candice
Lee, Annie Bond, and Tanner Feustel and others?? for assistance with
tissue collections. We thank Dr.~Fred Angelier for performing the
prolactin RIA.This research was supported by NSF IOS 1455960 to R.M.C.
and M.D.M. Author contributions: RMC designed the experiment. RMC and
MDM acquired funding. SHA contributed to the experimental design by
adding several time points. SHA, with the assistance of VF, AB, and RMC,
conducted the data collection. SHA and a team of trained undergraduate
students biopsy punched the hypothalami and lateral septum. ASL
extracted the RNA, created the libraries, and conducted all early
bioinformatic work. RMH conducted all later analyses of the genomic
data. RMH and SHA drafted the manuscript and all other authors provided
edits.

@@

\includegraphics[width=0.95\linewidth]{fig1-1}

\textbf{Figure 1. Experimental design.} A) We investigated nine
timepoints that spanned the majority of reproductive efforts in this
species. These time-points consisted of a ``control'' or non-parenting
state (from MacManes et al 2016 and Calisi et al.~2017), nest-building,
clutch initiation or the onset of incubation, clutch completion and
early incubation, mid-incubation, late incubation, initiation of
nestling care, early nestling care, and mid-nestling care. To teste
whether external or internal factors were regulating gene activity, we
also conducted a series of offspring removal (B) and replacement (C)
manipulations to test whether transcriptional changes were dependent
upon offspring presence or were regulated by an internal clock. (D) We
used t-Distributed Stochastic Neighbor Embedding (t-SNE) to reduce the
dimensionality of the transcriptomes from the hypothalamus, pituitary,
and gonads of male and female rock. Circles and triangles represent
female and male samples, respectively, and points are colored by
treatment.

\includegraphics[width=0.95\linewidth]{fig2-1}

\textbf{Figure 2. Hypothesis-driven and data-driven approaches to
identifying changes in gene expression.} Box-and-whisker plots show
changes in gene expression of serotonin receptor (HTR2C) in the
hypothalamus (A), prolactin (PRL) in the pituitary (B), and estrogen
receptor 1 (ESR1) in the gonads.Bar and statistics are shown for one
contrast of interest between treatment groups. Boxes are colored by
treatment. Volcano plots show the log-fold change (LFC) and -log10(FDR)
for all genes that are significantly different differentially between
the aforementioned treatment groups. Boxes are colored by the treatment
were gene expression was higher.

\includegraphics[width=0.95\linewidth]{fig3-1}

\textbf{Figure 3. Summary of differentially expressed genes.} Bar plots
show the total number of significantly differentially expressed genes
for all two-way comparisons in the hypothalamus (A), pituitary (B),
gonads (C) . Contrasts are made relative to the non-breeding controls,
the nest-building group, the previous stage, or the internal or external
control group. Bars are colored by the treatment where gene expression
was higher. A negative number of differentially expressed genes means
that many genes had increase expression in the reference group.

>>>>

% !TeX root = RJwrapper.tex
\title{The reproductive and parental care transcriptome of the rock dove}
\author{by Suzanne H. Austin \footnote{equal contribtion, currently at the
  University of Oregon}, Rayna M Harris \footnote{equal contribtion}, Andrew S. Lang, Victoria S. Farrar, April Booth, Tanner Feustel, Matthew D. MacManes, Rebecca M. Calisi \footnote{corresponding author}}

@@

\abstract{%
In animals that exhibit a parental care reproductive strategy,
successful rearing of offspring typically involves a shift from sexual
behaviors to more caring and nurturing ones. Decades of work have
highlighted the role of particular hormones in the maintenance of
parental care behavior, the majority of which are produced by a
biological system essential for reproduction, the
hypothalamic-pituitary-gonadal (HPG) axis. However, we know very little
about when and how the HPG axis transitions into parental care behaviors
in any vertebrate, and this knowledge is fundamentally important to our
understanding of the mechanisms mediating parental care. First, we
characterized the transcriptome of the HPG axis in both mothers and
fathers over 9 points of the parental care stage using a socially
monogamous and bi-parental model, the rock dove (Columba livia). Then,
we experimentally manipulated offspring presence at 7 different time
points using classic offspring replacement and removal manipulations to
uncover cause and effect of these transcriptional changes. We offer the
most complete characterization and understanding of changes in neural,
pituitary and gonadal transcription and translation ever reported in any
vertebrate over the course of parental care.
}

@@

Parental care is essential to many organisms' reproductive strategy and
predominates avian and mammalian groups (Clutton-Brock, 1991; Royle et
al., 2012). Depending on the species, reproductive and parental
behaviors may include territory establishment, courtship and mating,
nest-building, nest defense, incubation, and the brooding and
provisioning of young. The goal of these behaviors, either directly or
indirectly, is the successful production of offspring resulting in a
direct fitness benefit (Reviewed in Santos, E.S.A, 2012). Underlying
these behaviors are the anatomical and physiological changes that
facilitate the transition into and across reproductive and parenting
stages. These changes have fascinated biologists and inspired a
prodigious body of research. Much of this research has focused on
understanding reproduction and parental care from an endocrine and
neuroendocrine perspective (reviewed in Ondrasek 2016, Kelly 2020,
Wingfield, Silver, Buntin). Reproduction and parental care is largely
regulated by a set of three tissues: The Hypothalamus in the brain, the
Pituitary, and the Gonads (testes and ovaries). Together, these tissues
are often referred to as the `reproductive' axis or the HPG. To activate
reproduction, the HPG initiates an endocrine cascade starting with the
production and release of gonadotropin-releasing hormone (GnRH) in the
hypothalamus which, in turn, stimulates the anterior pituitary to
produce luteinizing hormone (LH) and follicle-stimulating hormone (FSH).
LH and FSH travel to the gonads via the bloodstream where they induce
gonadal growth and steroidogenesis of estradiol, testosterone and
progesterone). These sex steroids have well established roles in mating
behavior, nest-building, territory defense, and egg-laying in birds
(reviewed in Buntin). Another key hormone, prolactin, is produced in the
pituitary and is associated with the induction and maintenance of
behaviors such as incubation in birds and lactation (in doves and
mammals) (birds: reviewed in Angelier, Austin et al.~in prep, mammals:
Austin and Word). Other hormones like vasopressin and oxytocin are
related to affiliative behaviors (Yoshihara C, 2018) while dopamine,
serotonin aggressive behaviors or avoidance (Rilling, JK, 2013; Adams
DB, 2006). While these and many other hormones have been reported to
play integral roles in various reproductive and parental care behaviors,
we have yet to come close to conceptualizing the larger physiological
and genomic coordination of events that yield such behaviors.

To begin to address this larger goal, we investigated genomic changes in
the reproductive axis using the model of the rock dove (Columba livia).
This species is socially and genetically monogamous and has bi-parental
incubation and offspring care thereby making investigations of sexual
variation across reproduction possible. Doves and pigeons have been used
extensively to study the physiological mechanisms of reproduction and
parental care, including the proximate and ultimate mechanisms that
regulate associated behaviors (cite a bunch of new and old papers or
books). Our team has produced a series of papers with the goal of
illuminating how transcriptomic activity in the reproductive axis varied
by sex (MacManes et al.~2016), in response to a stressor (Calisi et
al.~2017, Lang et al.~2020), and following a hormonal manipulation
(Austin et al.~in prep). By using a genomic approach that spans the
reproductive timeline, we have the ability to elucidate a
transcriptional symphony that coordinates the anatomical, physiological
and behavioral changes associated with transitioning from non-breeding
to parenting in a vertebrate species. The aims of this research were
three-fold: Our first aim was to produce a high quality and reproducible
transcriptional characterization of the HPG over reproduction and
parental care in male and female rock doves. We were specifically
interested in understanding how gene activity in the HPG changed as
individuals transitioned from a sexually mature, non-reproductive state
into reproduction and parenting. To accomplish this aim, we investigated
nine time points that spanned the majority of reproductive efforts in
this species. These time points consisted of 1) a control or
non-parenting state (from MacManes et al 2016 and Calisi et al.~2017),
2) nest-building, 3) clutch initiation and onset of incubation, 4)
clutch completion and early incubation, 5) mid-incubation, 6) late
incubation, 7) initiation of nestling care, 8) early nestling care, and
9) mid-nestling care. Because the majority of time points included in
this study involved some form of direct offspring care, for simplicity,
we will discuss these as parental care rather than differentiating
between reproductive and parenting behaviors. However, it should be
understood that parental care remains only one component of a larger set
of reproductive behaviors. Our second aim focused on characterizing the
major patterns of transcriptional changes between the tissues, sexes,
and parental stages. We specifically investigate how gene expression in
the HPG changed across various transition points. For example, from
non-breeding to mating and nest-building behaviors, from nest-building
to egg-laying to incubation, from incubation to nestling care, and from
highly dependent nestlings (brooding and crop milk production) to older,
thermoregulatory independent nestlings that are fed a regurgitant of
seeds. Our third aim was to identify the drivers underlying changes in
transcription. Specifically, we tested whether external or internal
factors were regulating gene activity. We conducted a series of
offspring removal and replacement manipulations to test whether
transcriptional changes were dependent upon offspring presence or were
regulated by an internal clock.

Additionally, because these data are extensive, we took the opportunity
to provide them in an easily accessible form that can be used for
independent data exploration by the reader. Given the homology of genes
and gene networks shared between rock doves and other vertebrate
species, the findings from these RNA-seq experiments could be of
interest to scientists and clinicians studying diverse systems,
particularly in non-model organisms like our study species. Overall,
this study provides a deeper understanding of the interplay of genes
across parenting and has important implications for understanding
molecular processes that underlie behavioral transitions.

@@

\hypertarget{the-rock-dove-reproduction-and-parental-care-transcriptome}{%
\subsubsection{The rock dove reproduction and parental care
transcriptome}\label{the-rock-dove-reproduction-and-parental-care-transcriptome}}

We created the rock dove transcriptome by sequenced 987 RNA samples from
the hypothalamus, pituitary and gonads from the female and male of 166
rock dove pairs across (Fig. 1). These samples come from nine timepoints
that encapsulate important transitions in reproduction (Fig. 1A), four
offspring removal manipulations (Fig. 1B), and three offspring
replacement manipulations (Fig. 1C) to determine the influence of
internal mechanisms and external stimuli on gene expression. Using
T-distributed Stochastic Neighbor Embedding (tSNE) we show broad
patterns of variation due to treatment or sex in each tissue (Fig. 1D).
In the hypothalamus, tSNE dimension visually separates most of the
samples that were subject to an offspring manipulation from those from
naturally occurring parental stages. In the pituitary however, samples
cluster more by wether they were before incubation day 9 or after
incubation day 17. Neither the hypothalamus or pituitary show strong sex
differences at this level, but the gonads samples from very distinct,
sex-specific clusters.

\begin{Schunk}
\begin{figure}
\includegraphics{elife_v1_files/figure-latex/fig1-1} \caption[Experimental design]{Experimental design. A) We investigated nine timepoints that spanned the majority of reproductive efforts in this species. These timepoints consisted of a "control" or non-parenting state (from MacManes et al 2016 and Calisi et al. 2017), nest-building, clutch initiation or the onset of incubation, clutch completion and early incubation, mid-incubation, late incubation, initiation of nestling care, early nestling care, and mid-nestling care. To teste whether external or internal factors were regulating gene activity, we also conducted a series of offspring removal (B) and replacement (C) manipulations to test whether transcriptional changes were dependent upon offspring presence or were regulated by an internal clock. (D) We used t-Distributed Stochastic Neighbor Embedding (t-SNE) to reduce the dimensionality of the transcriptomes from the hypothalamus, pituitary, and gonads of male and female rock. Circles and triangles represent female and male samples, respectively, and points are colored by treatment.}\label{fig:fig1}
\end{figure}
\end{Schunk}

@@

@@

\hypertarget{tables}{%
\subsection{Tables}\label{tables}}

\begin{Schunk}
\begin{table}

\caption{\label{tab:table1}Changes in candidate gene expression. Significant chagns in gene expression are shown for X candidate genes. Positive and negative changes are denoted with (+) and (-) respectively for female (F) and male (M) hypothalamus (H), pituitary (P) and gonads (G). Comparisons are denoted in the column names. The first group is the reference group.}
\centering
\begin{tabular}[t]{l|l|l|l|l|l|l|l}
\hline
gene & bldg-lay & lay-inc3 & inc3-inc9 & inc9-inc17 & inc17-hatch & hatch-n5 & n5-n9\\
\hline
AR &  &  &  &  &  &  & \\
\hline
AVPR1B &  &  &  &  &  &  & \\
\hline
AVPR2 &  &  &  &  &  & FP(+) & \\
\hline
CRH &  &  &  &  &  &  & \\
\hline
CRHR1 &  &  &  &  &  &  & \\
\hline
DIO1 &  &  &  &  &  &  & \\
\hline
DIO2 &  & FG(-) &  &  &  &  & \\
\hline
DIO3 & FP(-) &  &  &  &  &  & \\
\hline
ESR2 &  &  &  &  &  &  & \\
\hline
FSHB &  &  &  &  &  &  & \\
\hline
FSHR &  &  &  &  &  &  & \\
\hline
GABRQ &  &  &  &  &  &  & \\
\hline
GALR1 &  &  &  &  &  &  & \\
\hline
GNRH1 &  &  &  &  &  &  & \\
\hline
GNRHR & FP(-) &  &  &  &  &  & \\
\hline
JAK2 &  &  &  &  &  &  & \\
\hline
LHCGR &  &  &  &  &  &  & \\
\hline
NPFFR1 &  &  &  &  &  &  & \\
\hline
NPVF &  &  &  &  &  &  & \\
\hline
NPY &  &  &  &  &  &  & \\
\hline
NR3C2 &  &  &  &  &  &  & \\
\hline
POMC &  &  &  &  &  &  & \\
\hline
PRL &  &  &  & FP(+); MP(+) &  &  & \\
\hline
PRLH & FP(-) & FP(+) &  & FP(+) &  &  & \\
\hline
PRLHR &  &  &  &  &  &  & \\
\hline
SERPINA4 &  &  &  &  &  &  & \\
\hline
VIPR1 &  &  &  &  &  &  & \\
\hline
\end{tabular}
\end{table}

\end{Schunk}

\begin{Schunk}
\begin{figure}
\includegraphics{elife_v1_files/figure-latex/fig2-1} \caption[Hypothesis-driven and data-driven approaches to identifying changes in gene expression.Box-and-whisker plots show changes in gene expression of serotonin receptor (HTR2C) in the hypothalamus (A), prolactin (PRL) in the pituitary (B), and estrogen receptor 1 (ESR1) in the gonads.Bar and statistics are shown for one contrast of interest between treatment groups]{Hypothesis-driven and data-driven approaches to identifying changes in gene expression.Box-and-whisker plots show changes in gene expression of serotonin receptor (HTR2C) in the hypothalamus (A), prolactin (PRL) in the pituitary (B), and estrogen receptor 1 (ESR1) in the gonads.Bar and statistics are shown for one contrast of interest between treatment groups. Boxes are colored by treatment. Volcano plots show the log-fold change (LFC) and -log10(FDR) for all genes that are significantly different differentially between the aforementioned treatment groups. Boxes are colored by the treatment were gene expression was higher.}\label{fig:fig2}
\end{figure}
\end{Schunk}

\begin{Schunk}
\begin{figure}
\includegraphics{elife_v1_files/figure-latex/fig3-1} \caption[Summary of differentially expressed genes]{Summary of differentially expressed genes. Bar plots show the total number of significantly differentially expressed genes for all two-way comparisons in the hypothalamus (A), pituitary (B), gonads (C) . Contrasts are made relative to the non-breeding controls, the nest-building group, the previous stage, or the internal or external control group. Bars are colored by the treatment where gene expression was higher. A negative number of differentially expressed genes means that many genes had increase expression in the reference group.}\label{fig:fig3}
\end{figure}
\end{Schunk}

@@

Thanks to Owen for discussion of data sonfification. Thanks to DIB lab
members for software and methods discussion (Titus, Taylor, Luiz,
Amanda, Alicia, Picaso). We thank Pacha for help with database and
Shiny. We thank Irene, Erica, Beth Krestoff and Tiffany Chen for
assistance in managing the aviary and collecting nest stage data. We
thank Candice Lee, Annie Bond, and Tanner Feustel and others?? for
assistance with tissue collections. We thank Dr.~Fred Angelier for
performing the prolactin RIA.This research was supported by NSF IOS
1455960 to R.M.C. and M.D.M. Author contributions: RMC designed the
experiment. RMC and MDM acquired funding. SHA contributed to the
experimental design by adding several time points. SHA, with the
assistance of VF, AB, and RMC, conducted the data collection. SHA and a
team of trained undergraduate students biopsy punched the hypothalami
and lateral septum. ASL extracted the RNA, created the libraries, and
conducted all early bioinformatic work. RMH conducted all later analyses
of the genomic data. RMH and SHA drafted the manuscript and all other
authors provided edits.
